\XX dubious ways of estimating $\rr$

### 2018-07-03 paper/old_drafts/draft_v2.tex
finally on git

\newcommand{\XX}{15 } % number of methods
\newcommand{\rr}{\ensuremath{\mathcal{R}_0}}

@@

%\tableofcontents

@@

\textbf{Point Estimation}
\begin{enumerate}
\item Survival function
  \begin{enumerate}
  \item Survival function
  \item Direct estimation of the survival function
  \item Direct parameter estimation
  \item Contact tracing
  \end{enumerate}
\item Compartment models
  \begin{enumerate}
  \item SIR model
  \item Harko's SIR model
  \item SEIR model
  \item Attack rate
  \item Amplifier model
  \item Sequential Bayes
  \item Next generation model
  \end{enumerate}
\item Exponential growth
  \begin{enumerate}
  \item Exponential growth
  \item Maximum likelihood of secondary infections
  \item Time dependent reproduction number
  \item Initial growth rate and final size
  \end{enumerate}
\item Estimation from networks
  \begin{enumerate}
  \item Branching process
  \item Agent-based models
  \end{enumerate}
  \end{enumerate}

@@

\textbf{Variance Estimation}
  \begin{enumerate}
  \item Inverse Hessian
  \item Sensitivity analysis
  \item Jackknife
  \item Delta method
  \end{enumerate}

@@

{$\mathcal{R}_t$} -- effective reproduction number
{$\beta$} -- the infection rate

@@

{$S(t)$} -- the number of susceptible individuals in a population at time $t$

{$I(t)$} -- the number of infected individuals in a population at time $t$

{$R(t)$} -- the number of recovered individuals in a population at time $t$

@@

\section{Details for the \XX Methods}
\label{sec:details}
\subsection{Survival function}
\label{sec:direct}
\subsubsection{Survival Function}
\label{sec:survival_fxn}
We describe the approach to the survival function as presented by \cite{Heffernan2005}.  They note that this is the most storied of the methods of calculating $\rr$.

Let $F(a)$ be the probability that a newly infected individual remains infectious for at least time $a$ (the survival probability).  Moreover, let $b(a)$ the average number of newly infected individuals that an infectious individual will produce per unit time when infected for total time $a$.  Then $\rr$ is derived in Equation (\ref{eq:r0_survivalfxn}),

@@

\com{this should be method \#1}

Estimating this quantity is hard, and it is due to this difficulty that the rest of these methods for estimating $\rr$ even exist.  This derivation of $\rr$ is not restricted to compartment models.

@@

\cite{fraser2004factors} have a derivation of $\rr$ related to the survival function, where $\beta (\tau)$ represents the infectiousness at time $\tau$ since the infection,

\begin{align*}
\rr = \int_0^\infty \beta(\tau) d\tau
\end{align*}

\com{move in with survival function.  These are \textit{mathematical quantities}}

@@

\cite{lipsitch2003} note that directly estimating $\rr$ in such a way may be difficult due to the available data.  They instead estimate $\rr$ using the mean serial interval, the time from onset of symptoms in an index case to the onset of symptoms in a subsequent case infected by the index patient in addition to the assumption of exponential growth of infected cases.

In the event, that th is estimate may be calculated, uncertainty estimates must be compounded, and to estimate $\rr$ alone, we only use data from the initial outbreak of the disease, so often we have a small sample size, resulting in large error bars.

%%%%%%%% 5
\subsubsection{Contact Tracing}
\label{sec:contact_tracing}
As $\rr$ is defined by \cite{anderson1992} ``the average number of secondary infections produced when one infected individual is introduced into a host population where everyone is susceptible,'' it seems reasonable to actually observe the number of secondary infections produced by an initial infector.  This is generally impractical, with seriological information, we could conceivably recreate the transmission of the disease throughout a population.  Even in this case, however, $\rr$ it is unlikely, for one, to have a completely susceptible population.  Moreover, this would implicitly be conditioning on Patient 0's contacts which may not reflect those of the rest of the population.  We write the calculation for $\rr$ under contact tracing in Equation \ref{eq:r0_contacttracing}, where $X_0$ is a random variable denoting the number of individuals in the population person 0 infects.  Here $S_0=N$ indicates that the initial number of susceptibles is $N$, the size of the population.

\cite{eames2003} describe contact tracing as an ``extreme form of targeted control, where the potential next-generation cases are the primary focus,'' i.e. one focuses on treating the contacts of an infected individual rather than applying random treatment.  They note that this method has ``proved to be a highly successful strategy when the number of infectious cases is low'' and ``is the standard tool for eliminating minor outbreaks in the latter stages of disease eradication.''

\com{should this be split into two methods?}

@@

@@

In practice, $\rr$ is often calculated through fitted compartment models.  An example of this is the SIR model.  First introduced to epidemiology by  \cite{Kermack700}, the SIR model governs how individuals transition from \textbf{S}usceptible, \textbf{I}nfected, and \textbf{R}ecovered compartments.  The model is represented graphically in Figure \ref{fig::sir}.  In this model, once a person is recovered she cannot have the disease again.

\begin{figure}[h]
\centering
\begin{tikzpicture}[node distance=3cm,auto,>=latex',every node/.append style={align=center}]
    \node [int,  fill = white!50!green] (a)              {$S$};
    \node [int,  fill = white!50!green]           (c) [right of=a] {$I$};
    \node [int,  fill = white!50!green] (e) [right of=c] {$R$};
    \path[->, auto=false] (a) edge node {} (c)
                          (c) edge node {} (e) ;

\end{tikzpicture}
\caption{Depiction of a SIR model.  One can only get the disease once in this model.}\label{fig::sir}
\end{figure}

Here, $N$ is the number of individuals in the system, and  $\beta$ and $\gamma$ are epidemiological parameters.  We have that $\beta$ is the infection rate and $\gamma$ is the inverse of the recovery rate.  The model is represented through the ODEs in \ref{eq::sir}.

A key point in the SIR model and other CMs is the assumption of homogeneity.  This means that ``any infected individual has a probability of infecting any susceptible individual that is reasonably well approximated by the average'' \cite{jonesjh2007}

\begin{align}
\systeme{\frac{dS}{dt} = -\frac{\beta I S}{N}, \frac{dI}{dt} = \frac{\beta I S}{N} - \gamma I, \frac{dR}{dt} = \gamma I}. \label{eq::sir}
\end{align}

@@

An epidemic occurs if 
\begin{align*}
\frac{\beta S I}{N}  - \gamma I > 0 ,
\end{align*}

e.g. the rate of infection is greater than the rate of recovery.  So as long as $\frac{S}{N} \approx 1$ then an epidemic will occur if $\rr := \frac{\beta}{\gamma} > 1$.

@@

There are many ways to estimate the parameters in this model.  We detail a few of them here.

\textbf{Fitting Methods}\label{fitting-methods}

\textbf{E{[}\(\rr\){]}}:

\begin{enumerate}
\item
  Least Squares (\(\beta\), \(\gamma\)) (LS)
\item
  Reparametrized Least Squares (\(\rr\), \(\gamma\)) (ReLS)
\item
  Linear Model Approximation (LMA)
\item
  Linear Model Approximation, all time points (LMAT)
\item
  Max of data (Max)
\item
  Smooth max (spline) (SMax)
\item
  Incidence-Prevalence Ratio (IPR)
\end{enumerate}

\textbf{Least Squares ($\beta$, $\gamma$)}\label{least-squares-beta-gamma}

@@

\textbf{Reparametrized Least Squares ($\rr$, $\gamma$)}\label{reparametrized-least-squares-rux5f0-gamma}

We Reparametrized the the ODES directly with \(\rr\) and \(\gamma\),

@@

\textbf{Linear Model Approximation (degree
10)}\label{linear-model-approximation-degree-10}

We fit a linear polynomial of \(t\) with degree \(K= 10\) model to \(X\)
and \(Y\) using least squares to find the coefficients (\(\hat{x}_k\),
\(\hat{y}_k\)),

@@

\textbf{Linear Model Approximation, All Time Points (degree
10)}\label{linear-model-approximation-all-time-points-degree-10}

We fit a linear polynomial of \(t\) with degree \(K= 10\) model to \(X\)
and \(Y\) as above, with a slight modification in how we calculate
\(\rr\),
\[\rr = \frac{1}{\# \text{ Obs}}\sum_t \frac{\hat{X}^\prime(t)}{ \hat{X}^\prime(t) + \hat{Y}^\prime(t)} \cdot \frac{N}{X(0)} \]

\textbf{Max of Data}\label{max-of-data}

We note that \(Y^\prime(t) > 0\) when \(\rr < \frac{S(t)}{N}\). The max
of \(Y\) occurs when \(Y^\prime = 0\) and hence
\[\rr = \frac{N}{X_{obs}(t^*)},\] where
\(t^* = \text{arg max}_{t} Y_{obs}(t)\).

\textbf{Smooth Maximum of Data}\label{smooth-maximum-of-data}

@@

\textbf{Incidence to Prevalence
Ratio}\label{incidence-to-prevalence-ratio}
In terms of data from the SIR model, incidence $J(t) \approx -(X(t+1) - X(t))$, and the IPR$(t) = \frac{J(t)}{Y(t)}$. The actual reproduction number, $R_a(t) = \text{IPR}(t)\cdot D$ where $D = 1 /\gamma$.  This method assumes that we have some prior knowledge about $\gamma$.  Thus we use as our estimate,

\textbf{Smoothed Incidence to Prevalence Ratio}
We use the same method as above, but first fit splines with 4 degrees of freedom, $\hat{X}(t)$ and $\hat{Y}(t)$, , to fit to $X_{obs}$ and $Y_{obs}$, respectively.  Then  $J(t) \approx -(\hat{X}(t+1) - \hat{X}(t))$, and the IPR$(t) = \frac{J(t)}{\hat{Y}(t)}$.  Then
\begin{align*}
\rr &= R_a(0).
\end{align*}

\subsubsection{Harko}
Recently, \citep{harko2014exact} were able to reduce the SIR (XYZ) equation, with the following set of ODEs

@@

If we assume $\epsilon_t \overset{iid}{\sim}N(0, \sigma^2)$, then we can directly estimate $\rr/N$ through linear regression and also an expression for the variance.

\subsubsection{SEIR Model}
\label{sec:seir-model}

The SEIR model is a common adaptation of the SIR compartment model \com{add references that use this method}.  The ``E'' compartment stands for ``exposed'' and represents the stage where individuals are infected but not yet infectious.  The model is described by the following set of equations, as described by \cite{cintronarias2009}.  This particular model here does not include birth and death rates but the model can be adapted to do so,
\begin{align*}
  \frac{dS}{dt} &= - \beta SI \\
  \frac{dE}{dt} &= \beta SI  - \alpha E\\
  \frac{dI}{dt} &= \alpha E - \gamma I \\
  \frac{dR}{dt} &= \gamma I.
\end{align*}

Then $\rr = \beta / \gamma$, just as in the SIR model.  As in the SIR model, $\beta$ and $\gamma$ have the same interpretation and $\alpha$  The methods used to estimate the parameters $\beta$ and $\gamma$ are of the same class used to estimate those parameters in the SIR model.  The important difference between this method and the SIR model is the assumption of a 4th class, the exposed compartment.  Typically, this method will result in ...\com{figure out}... than the SIR method.  Homogeneous mixing of the compartments is assumed.

\subsubsection{Attack Rate}

@@

\subsubsection{Amplifier Model}
\cite{blower2004} discuss the case where a disease may have different strains and consequently different reproduction numbers.  They introduce a multistrain mathematical model called the amplifier model to deal with this case.

This model is based on the the SLIR compartment model, where $L$ stands for latently infected individuals and $T$ for diseased individuals.  There is a set of ODEs for each strain of the disease with different parameters for recovery, transmission rate, and drug resistance of the disease.  Additionally, there is mixing of the different strains as ``the amplifier model also allows for immigrating and emigration of individuals of all types, as well as reinfection of latently infected individuals'' \citep{blower2004}.  The details may be found in their manuscript.  Ultimately, their derivation for $\rr^{(i)}$ where $i$ stands for the $i$th strain is
\begin{align*}
\rr^{(i)} = S^* \frac{ ( \beta_i^T + \beta_i^L)\nu_i + \beta_i^T \mu_i^L}{(\nu_i + \mu_i^L)(c_i + k_{i,i+1} + \mu_i^T)}.
\end{align*}
Here, $S^*$ is the number of susceptibles individuals in the disease-free equilibrium in the absence of immigration or emigration; $\beta_i^L$ and $\beta_i^T$ are strain-specific transmission rates for the latent, and infected classes, respectively; $\nu_i$ denotes the progression rate from latently infected to disease for strain $i$; $\mu_i^L$ and $\mu_i^{T}$ are death rates for the latent and infected classes, respectively; $c_i$ is the cure rate; and $k_{i, i+1}$ is a parameter associated with the degree of amplification of drug resistance.  They note that ``each reproduction number is the product of average number of secondary infections caused per unit time, the aver time a case remains infectious and the probability that an infected individual develops disease.''

In the amplifier model, $\rr$ was derived from using the stability point (disease free equilibrium state in the absence of immigration or emigration) of the system by solving the characteristic polynomial for the eigenvalues

\subsubsection{Sequential Bayes}\label{sec:seqbayes}

Posited by \cite{bettencourt2008} and summarized in \cite{obadia2012r0}, this method uses a Bayesian approach to an approximation to the classic SIR model.  The approximated SIR model is that the incidence at $t+1$, $N(t+1)$ is, with $\gamma$ as the  average inverse of the infectious period and $R$ as the effective reproduction number, which we take here to be $\rr$.

In order to estimate $\rr$, we must have some idea about the serial interval.
\begin{align*}
N(t+1)  \sim Poisson( N(t) \exp \left \{  \gamma (R-1)\right \})
\end{align*}
Then, the posterior distribution of $\rr$ given the previous days' incidences is
\begin{align*}
  P(\rr | N_0, \dots, N_{t+1}) = \frac{P(N_{t+1} | \rr, N_0, \dots, N_t)P(\rr| N_0, \dots, N_t)}{P(N_0, \dots, N_{t+1})}.
\end{align*}
This method is sequential in that the prior distribution for $\rr$ comes from the previous day.  The initial prior for $\rr$ is assumed to be flat.  This method results in a posterior distribution from which credible intervals may be obtained.  This method assumes, initial growth in incidence to be exponential, and homogeneous mixing of populations as with any compartment model.

\subsubsection{Next Generation Model}
\label{sec:ngm}

@@

This method has the assumptions of a basic compartment model: homogeneous mixing among populations and the law of mass action, i.e., that the number of infected individuals are proportional to the number of infected individuals at the previous step.  The estimations for $\rr$ from the previously mentioned compartment models \com{list those models} may also be derived using this method.  Additionally, this method is advantageous in that the matrix $\Sigma$ has an intuitive meaning.

@@

\subsection{Exponential Growth}\label{sec:exp-growth}

\subsubsection{Exponential Growth}
\label{sec:expgrowth}
\cite{wallinga2007generation} report that $R$ and hence $\rr$ may calculated by using the fact that infection ``counts increase exponentially in the initial phase of an epidemic.''  We need to know $r$, the \textit{per capita} change in the number of new cases per unit of time and $T_c$ the serial interval.

@@

\begin{align}\label{eq:lotka}
\rr = \exp{(r T_c)}
\end{align}
or its first order approximation

\begin{align}\label{eq:anderson}
\rr = 1 + rT_c.
\end{align}

>>>>

\newcommand{\XX}{21 } % number of methods
\newcommand{\rr}{\ensuremath{\mathcal{R}_0}}

@@

\tableofcontents

@@

\textbf{Point Estimation}
\begin{enumerate}
\item Survival function
  \begin{enumerate}
  \item Survival function (SF)
  \item Direct estimation of the survival function (DESF)
  \item Direct parameter estimation (DPE)

  \end{enumerate}
\item Compartment models 
  \begin{enumerate}
  \item SIR model
    \begin{itemize}
    \item Least Squares (SIR LS)
    \item Reparametrized Least Squares (SIR ReLS)
    \item Linear Model Approximation (SIR LMA)
    \item Linear Model Approximation, all time points (SIR LMAT)
    \item Max of Data (SIR Max)
    \item Smooth max (SIR SMax)
    \item Incidence-Prevalence Ratio (SIR IPR)
    \item Smooth Incidence-Prevalence Ratio (SIR SIPR)
    \item Harko's SIR model (HSIR)
    \item Attack rate (SIR AR)
    \item Sequential Bayes (SIR SB)
    \end{itemize}
  \item SEIR model (SEIR)
  \item Amplifier model (Amp)
  \item Next generation model (NGM)
  \end{enumerate}
\item Exponential growth 
  \begin{enumerate}
  \item Exponential growth (EG)
  \item Maximum likelihood of secondary infections (MLSI)
  \item Time dependent reproduction number (TDR)
  \item Initial growth rate and final size (IGRFS)
  \end{enumerate}
\item Estimation from networks 
  \begin{enumerate}
  \item Contact tracing (CT)
  \item Branching process (BP)
  \item Agent-based models (ABM)
  \end{enumerate}
  \end{enumerate}

@@

\textbf{Variance Estimation}
\begin{enumerate}
\item Delta method
  \item Sensitivity analysis
  \end{enumerate}

@@

{$\mathcal{R}_t$} -- effective reproduction number

{$\beta$} -- the infection rate

@@

{$X(t)$} -- the number of susceptible individuals in a population at time $t$

{$Y(t)$} -- the number of infected individuals in a population at time $t$

{$Z(t)$} -- the number of recovered individuals in a population at time $t$

@@

\section{Details for the \XX Methods}
\label{sec:details}
\subsection{Survival function}
\label{sec:direct}

The survival function is intimately with the theory of \rr, which originates from the field of demography in the late 1800s \citep{dietz1993estimation}.  Here, we have
\begin{align*}
\rr = \int_0^\infty p(a) \beta(a) da
\end{align*}
where $p(a)$ denotes the probability of a woman surviving age $a$ and $\beta(a)$ the rate of giving birth to a  girl for an individual of age $a$.  This interpretation also explains the name of \rr, which was imported to the field of epidemiology by MacDonald and Smith as reported by \cite{dietz1993estimation}.

This is the chronologically first and in many ways, the most direct method to calculate \rr.  However, this model also seems to be the most difficult to estimate.  We describe four methods attempting to estimate \rr from the survival function:  the survival function alone, direct estimation, and direct parameter estimation.

\subsubsection{Survival Function}
\label{sec:survival_fxn}
The first method, is not surprisingly, the survival function itself as presented by \cite{Heffernan2005}.  We change the notation from the above survival function to conform to the standards of epidemiology.

Let $F(a)$ be the probability that a newly infected individual remains infectious for at least time $a$ (the survival probability,$p(a)$ in the previous definition).  Moreover, let $b(a)$ $(\beta (a)$ in the previous definition), the average number of newly infected individuals that an infectious individual will produce per unit time when infected for total time $a$.  Then $\rr$ is derived in Equation (\ref{eq:r0_survivalfxn}),

@@

This method presents one of the unique difficulties of estimating \rr: that of reconciling generations and time.  On the other hand, this method is flexible in that this derivation of $\rr$ is not restricted to compartment models.

Estimating this quantity in Equation \ref{eq:r0_survivalfxn} is hard and rarely attempted as $b(a)$ is particularly difficult to estimate, and it is due to this difficulty that the rest of these methods for estimating $\rr$ even exist.

@@

\cite{fraser2004factors} have a derivation of $\rr$ related to the survival function based on the Van Forester equations,
\begin{align*}
  \frac{\partial Y(t, \tau)}{\partial t} +   \frac{\partial Y(t, \tau)}{\partial \tau} = 0,
\end{align*}
where $Y(t,\tau)$ is the number of people at time $t$ who were infected time $\tau$ ago.  $\rr$  is then,
\begin{align*}
\rr = \int_0^\infty \beta(\tau) d\tau,
\end{align*}
 where $\beta (\tau)$ represents the infectiousness at time $\tau$ since the infection.  Again, this model is more mathematical in nature than practical as $\beta (\tau)$ is difficult to estimate.

@@

\cite{lipsitch2003} note that directly estimating $\rr$ in such a way may be difficult due to the available data.  They instead estimate $\rr$ using the mean serial interval, $\omega$, the time from onset of symptoms in an index case to the onset of symptoms in a subsequent case infected by the index patient in addition to the assumption of exponential growth of infected cases.

In the event, that the is estimate may be calculated, uncertainty estimates must be compounded, and to estimate $\rr$ alone, we only use data from the initial outbreak of the disease, so often we have a small sample size, resulting in large confidence sets.

@@

A large and the most common set of models used estimate $\mathcal{R}$ is that of compartment models (CMs).  As adapted from \cite{daley2001epidemic}, compartment models consist of the following.  CMs model how objects move from one discrete state to another over time and provide a high-level model of how diseases, for instance, may be spread throughout a population.  Within the CM framework, we make four essential assumptions,
\begin{enumerate}
\item The compartments are discrete and have no overlap.
\item The transition of objects among compartments is described by a set of equations.
\item The populations mix homogeneously
  \item The number of objects in each compartment at time $t=0$ is known
  \end{enumerate}

Another common assumption for compartment models, is the law of mass action, a property borrowed from chemistry which says that the mass of the product is proportional to the mass of the reactants.  \cite{anderson1992} adapt the law of mass action to epidemiology saying, ``the net rate of spread of infection is assumed to be proportional to the product of the susceptible people times the density of the infectious individuals.''  Therefore, the flow of objects from one compartment to another may be dependent on the percentage of objects within one or many of the compartments.  The assumption of mass action holds particularly true for disease modelling as typically we assume infections at the next state are proportional to both the susceptible and infected populations at the current state.

For disease modelling, all CMs fall into the SI-framework, meaning that there is a susceptible population that may become infected.  Thus, we require there be susceptible and infected compartments but also allow additional compartments such as the recovered state, an exposed latent state, and more.  CMs may either work with continuous or discrete time and \cite{getz2006} remark that there is no preference between the two.

In this section, we discuss 14 different CMs used to estimate $\mathcal{R}$.  Different CMs yield different mathematical derivations of \rr.  That is, for example, an SIR model with non-constant population will yield a different mathematical formula for \rr than the SEIR model with non-constant population.  The only difference in these two models is that the SEIR model.  That said, even within the same compartment model framework, we can estimate \rr in a variety of ways.  In fact, we describe 11 methods to estimate \rr, within the SIR framework!  Most of these methods are descriptions to estimate parameters within the SIR model and may be extended to other CM frameworks such as the SEIR model.  The final method described in this section, the next generation operator (NGM) is a general method to derive \rr from almost any CM.

@@

We begin with the most popular CM, the SIR model.   First introduced to epidemiology by  \cite{Kermack700}, the SIR model governs how individuals transition from \textbf{S}usceptible, \textbf{I}nfected, and \textbf{R}ecovered compartments.  The model is represented graphically in Figure \ref{fig::sir}.  In this model, once a person is recovered she cannot have the disease again.

\begin{figure}[h]
\centering
\begin{tikzpicture}[node distance=3cm,auto,>=latex',every node/.append style={align=center}]
    \node [int,  fill = white!70!blue] (a)              {$S$};
    \node [int,  fill = white!70!blue]           (c) [right of=a] {$I$};
    \node [int,  fill = white!70!blue] (e) [right of=c] {$R$};
    \path[->, auto=false] (a) edge node {} (c)
                          (c) edge node {} (e) ;

\end{tikzpicture}
\caption{Depiction of a SIR model.  One can only get the disease once in this model.}\label{fig::sir}
\end{figure}
Here, $N$ is the number of individuals in the system.  As we see no arrows pointing out of the nodes in Figure \ref{fig::sir}, then $N$ is constant throughout time.  Simple adaptations of the SIR model can include birth and death rates, which will correspondingly change the derivation of $\rr$ (for more information, see Section \ref{sec:ngm}). The parameters  $\beta$ and $\gamma$ have epidemiological meaning.  Here, $\beta$ is the average infection rate and $\gamma$ is the inverse of the average recovery rate.  The model is represented through the ODEs in Equation \ref{eq::sir}.  Throughout, we use $X$, $Y$, and $Z$ for stand-ins for the ODEs for $S$, $I$, and $R$, respectively to avoid confusion.
\begin{align}
\systeme{\frac{dX}{dt} = -\frac{\beta XY}{N}, \frac{dY}{dt} = \frac{\beta XY}{N} - \gamma Y, \frac{dZ}{dt} = \gamma Z}. \label{eq::sir}
\end{align}

@@

An epidemic occurs if 
\begin{align*}
  \frac{dY}{dt} &> 0 \\
\frac{\beta S I}{N}  - \gamma I &> 0 ,
\end{align*}

that is,  the rate of new infections is greater than the rate of recovery.  So as long as $\frac{S}{N} \approx 1$ then an epidemic will occur if
\begin{align}\label{eq:sir_r0}
  \rr = \frac{\beta}{\gamma} > 1.
  \end{align}

@@

There are numerous ways to estimate the parameters in this model.  We detail 11 of them here.

\subsubsection{Least Squares ($\beta$, $\gamma$) (SIR LS)}\label{least-squares-beta-gamma}

@@

We can use gradient descent or another solver to find estimates for these parameters.  If we assume $Y= f(t) + \epsilon_t$ with $\epsilon_t \overset{iid}{\sim} N(0, \sigma^2)$ where $f(t) = \int_0^t \frac{dY}{dt} dt$, then this this method is equivalent to finding the MLE of $\beta$ and $\gamma$.  However, we note that the assumption of homoskedasticity seems likely to be violated in this model.

\subsubsection{Reparametrized Least Squares ($\rr$, $\gamma$) ( SIR ReLS)}\label{reparametrized-least-squares-rux5f0-gamma}

We reparametrized the the ODES in Equation \eqref{eq::sir} directly with \(\rr\) and \(\gamma\), using the relation $\rr = \frac{\beta}{\gamma}$.

@@

\subsubsection{Linear Model Approximation (SIR LMA)}\label{linear-model-approximation-degree-10}

The idea that an SIR model may be well approximated by a linear model is described by \cite{chang2017}.  We borrow this idea here and use the ODES in Equation \eqref{eq::sir} to relate our estimated linear model to that of $\rr$.

Specifically, we fit a linear polynomial of \(t\) with degree \(K= 10\) model to \(X\)
and \(Y\) using least squares to find the coefficients (\(\hat{x}_k\),
\(\hat{y}_k\)),

@@

Like in the previous cases, we are assuming the true models are

\begin{align*}
X(t) &= \sum_{k=0}^K x_k t^k + \epsilon_t\\
  {Y}(t) &= \sum_{k=0}^K y_k t^k + \epsilon_t,
\end{align*}
with $\epsilon_t \overset{iid}{\sim} N(0, \sigma^2)$.  Here, $K=10$ is rather arbitrary and should be selected using some criterion such as AIC in addition with the knowledge of how many data points are available to the user.

\subsubsection{Linear Model Approximation, All Time Points (SIR LMAT)}\label{linear-model-approximation-all-time-points-degree-10}

We fit a linear polynomial of \(t\) with degree \(K= 10\) model to \(X\)
and \(Y\) as above, with a slight modification in how we calculate
\(\rr\),
\[\rr = \frac{1}{\# \text{ Obs}}\sum_t \frac{\hat{X}^\prime(t)}{ \hat{X}^\prime(t) + \hat{Y}^\prime(t)} \cdot \frac{N}{X(0)} \]
The idea is that $\frac{\mathcal{X}^\prime}{X^\prime + Y^\prime}$ is constant in $t$, but due to our approximations with the linear model, this is no longer the case.  Here, we average over the different possible values of $\rr(t)$.

\subsubsection{Max of Data (SIR Max)}\label{max-of-data}

Under the SIR model there is an absolute maximum of infected individuals $Y$.  We note that \(Y^\prime(t) = \frac{\beta XY}{N} - \gamma Y > 0\) when \(\rr = \frac{\beta}{\gamma} < \frac{X(t)}{N}\). The max
of \(Y\) occurs when \(Y^\prime = 0\) and hence
\[\rr = \frac{N}{X_{obs}(t^*)},\] where
\(t^* = \text{arg max}_{t} Y_{obs}(t)\).

\subsubsection{Smooth Maximum of Data (SIR SMax)}\label{smooth-maximum-of-data}

@@

\subsubsection{Incidence to Prevalence
Ratio (SIR IPR)}\label{incidence-to-prevalence-ratio}
In terms of data from the SIR model, incidence $J(t) \approx -(X(t+1) - X(t))$, and the IPR$(t) = \frac{J(t)}{Y(t)}$. The actual reproduction number, $R_a(t) = \text{IPR}(t)\cdot D$ where $D = 1 /\gamma$.  This method assumes that we have some prior knowledge about $\gamma$.  Thus we use as our estimate,
\begin{align*}
\rr &= R_a(0).
\end{align*}

Here we assume that the time step is small enough to approximate the incidence.

\subsubsection{Smoothed Incidence to Prevalence Ratio (SIR SIPR)}
We use the same method as above, but first fit splines with 4 degrees of freedom, $\hat{X}(t)$ and $\hat{Y}(t)$, , to fit to $X_{obs}$ and $Y_{obs}$, respectively.  Then  $J(t) \approx -(\hat{X}(t+1) - \hat{X}(t))$, and the IPR$(t) = \frac{J(t)}{\hat{Y}(t)}$.  Then
\begin{align*}
\rr &= R_a(0).
\end{align*}

%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Harko (HSIR)}
Recently, \citep{harko2014exact} were able to reduce the SIR equation, with the following set of ODEs (reparametrized from Equation \eqref{eq::sir})

@@

If we assume $\epsilon_t \overset{iid}{\sim}N(0, \sigma^2)$, then we can directly estimate $\rr/N$ through linear regression and also an expression for the variance.  This method greatly simplifies the computational effort needed to estimate $\rr$ as it does not require any numerical integration.

\subsubsection{Attack Rate (SIR AR)}

@@

\subsubsection{Sequential Bayes (SIR SB)}\label{sec:seqbayes}

Posited by \cite{bettencourt2008} and summarized in \cite{obadia2012r0}, this method uses a Bayesian approach to an approximation to the classic SIR model.  The approximated SIR model is that the incidence at $t+1$, $N(t+1)$ is from a Poisson distribution, with $\gamma$ as the  average inverse of the infectious period and $R$ as the effective reproduction number, which we take here to be $\rr$.

In order to estimate $\rr$, we must have some idea about the serial interval, $\omega$.
\begin{align*}
N(t+1)  \sim Poisson( N(t) \exp \left \{  \gamma (R-1)\right \})
\end{align*}
Then, the posterior distribution of $\rr$ given the previous days' incidences is
\begin{align*}
  P(\rr | N_0, \dots, N_{t+1}) = \frac{P(N_{t+1} | \rr, N_0, \dots, N_t)P(\rr| N_0, \dots, N_t)}{P(N_0, \dots, N_{t+1})}.
\end{align*}
This method is sequential in that the prior distribution for $\rr$ comes from the previous day.  The initial prior for $\rr$ is assumed to be flat.  This method results in a posterior distribution from which credible intervals may be obtained.  This method assumes, initial growth in incidence to be exponential, and homogeneous mixing of populations as with any compartment model.

\subsubsection{SEIR Model (SEIR)}
\label{sec:seir-model}

The SEIR model is a common adaptation of the SIR compartment model \com{add references that use this method}.  The ``E'' compartment stands for ``exposed'' and represents the stage where individuals are infected but not yet infectious.  The model is described by the following set of equations, as described by \cite{cintronarias2009}.  This particular model here does not include birth and death rates but the model can be adapted to do so,
\begin{align*}
  \frac{dX}{dt} &= - \frac{\beta XY}{N} \\
  \frac{dE}{dt} &= \frac{\beta XY}{N}  - \alpha E\\
  \frac{dY}{dt} &= \alpha E - \gamma Y \\
  \frac{dZ}{dt} &= \gamma Y.
\end{align*}

Then $\rr = \beta / \gamma$, just as in the SIR model.  As in the SIR model, $\beta$ and $\gamma$ have the same interpretation and $\alpha$  The methods used to estimate the parameters $\beta$ and $\gamma$ are of the same class used to estimate those parameters in the SIR model.  The important difference between this method and the SIR model is the assumption of a 4th class, the exposed compartment.  Even though the derivation for $\rr$ is the same for both the SIR and SEIR model, this will still result in different results for estimations of \rr, even if the same method for estimation of parameters is used.  Typically, some form of least squares is used to estimate parameters in this model. We demonstrate this in Section \com{results}.

\subsubsection{Amplifier Model (Amp)}
\cite{blower2004} discuss the case where a disease may have different strains and consequently different reproduction numbers.  They introduce a multistrain mathematical model called the amplifier model to deal with this case.

This model is based on the the SLIR compartment model, where $L$ stands for latently infected individuals and $T$ for diseased individuals.  There is a set of ODEs for each strain of the disease with different parameters for recovery, transmission rate, and drug resistance of the disease.  Additionally, there is mixing of the different strains as ``the amplifier model also allows for immigrating and emigration of individuals of all types, as well as reinfection of latently infected individuals'' \citep{blower2004}.  The details may be found in their manuscript.  Ultimately, their derivation for $\rr^{(i)}$ where $i$ stands for the $i$th strain is
\begin{align*}
\rr^{(i)} = X^* \frac{ ( \beta_i^T + \beta_i^L)\nu_i + \beta_i^T \mu_i^L}{(\nu_i + \mu_i^L)(c_i + k_{i,i+1} + \mu_i^T)}.
\end{align*}
Here, $X^*$ is the number of susceptibles individuals in the disease-free equilibrium in the absence of immigration or emigration; $\beta_i^L$ and $\beta_i^T$ are strain-specific transmission rates for the latent, and infected classes, respectively; $\nu_i$ denotes the progression rate from latently infected to disease for strain $i$; $\mu_i^L$ and $\mu_i^{T}$ are death rates for the latent and infected classes, respectively; $c_i$ is the cure rate; and $k_{i, i+1}$ is a parameter associated with the degree of amplification of drug resistance.  They note that ``each reproduction number is the product of average number of secondary infections caused per unit time, the aver time a case remains infectious and the probability that an infected individual develops disease.''

In the amplifier model, $\rr$ was derived from using the stability point (disease free equilibrium state in the absence of immigration or emigration) of the system by solving the characteristic polynomial for the eigenvalues.  This derivation is explained in more depth in Section \ref{sec:ngm}.

We show this model as opposed to other available CMs because it shows not only how complicated CMs can become, but researchers approach the assumption of homogeneity, which is often suspect.  In this case, we have strain-specific transmission rates, so we are essentially stratifying compartments into different sets.  Very complicated models have arisen from stratification of compartments, which leads to the approach of ABMs.

\subsubsection{Next Generation Model (NGM)}
\label{sec:ngm}

@@

This method has the assumptions of a basic compartment model: homogeneous mixing among populations and the law of mass action, i.e., that the number of infected individuals are proportional to the number of infected individuals at the previous step.  The estimations for $\rr$ from the previously mentioned compartment models may also be derived using this method.  Additionally, this method is advantageous in that the matrix $\Sigma$ has an intuitive meaning.

Although, this method is useful for deriving an expression of $\rr$, it is unclear how to actually estimate it from these expressions.  Again, least squares is typically used to estimate such parameters.

@@

\subsection{Exponential Growth}\label{sec:exp-growth}
Typically, CMs expressed through differential equations have no known analytical solutions, which makes deriving an expression for $\rr$ difficult.  One simplification that is often introduced to the SI framework is that of exponential growth, or that the number of infections in the beginning of an epidemic mimics that of an exponential curve.  This simplified expression has allowed for many derivations of $\rr$.

\subsubsection{Exponential Growth (EG)}
\label{sec:expgrowth}
\cite{wallinga2007generation} report that $\mathcal{R}_t$ and hence $\rr$ may calculated by using the fact that infection ``counts increase exponentially in the initial phase of an epidemic.''  We need to know $r$, the \textit{per capita} change in the number of new cases per unit of time and $\omega$ the serial interval.

@@

\begin{align}\label{eq:lotka}
\rr = \exp{(r \omega)}
\end{align}
or its first order approximation

\begin{align}\label{eq:anderson}
\rr = 1 + r \omega.
\end{align}

Equation \eqref{eq:lotka} is derived from a demographic view using the Lotka-Euler survival equations which come from the fields of demography, ecology, and evolutionary biology, whereas Equation \eqref{eq:anderson}is derived through an epidemiologist.  These estimates for $\rr$ between Equation \eqref{eq:lotka} and Equation \eqref{eq:anderson} can be strikingly different.
